Does Anybody Really Know What Time It Is?\\ Automating the Extraction of Date Scalars

### 2015-07-07 futurework.tex

While we are satisfied with the results so far, we found a number of date formats that we were not able to handle with these techniques. Many columns contained multiple formats, and we would like to recognize this situation and generate predicate calculations to increase the accuracy of the results and thereby make the results even more seamless for the user. Other columns contain date ranges, which we would like to handle by generating multiple calculations, possibly by combining regular expressions with date parsing.

In this paper, we have only considered the parsing of strings, but dates are often formatted as integers (\eg \texttt{201507016}). It is significantly faster to decompose integers into date parts using arithmetic operations (\eg \texttt{mod} and \texttt{div}) than by using slower, locale-sensitive string parsing functions. Moreover, the number of possible formats is low enough that we may be able to enumerate them. Timestamps are another form of numeric date for which we intend to automate the preparation.

In the course of our research, we have also identified a number of date part variants (\eg ordinal dates, four-letter month abbreviations, alternate meridian markers and postfix quarter syntax) that we would like to commend to the ICU project and maybe provide implementations for.

>>>>

While we are satisfied with the results so far, we found a number of column that we were not able to process with these techniques. Many columns contained multiple date formats, and we would like to recognize this situation and generate predicate-based calculations to increase the accuracy of the results and thereby make the experience even more seamless for the user. Other columns contain date ranges, which we would like to handle by generating multiple calculations, possibly by combining regular expressions with date parsing.

In this paper, we have only considered the parsing of strings, but dates are often formatted as integers (\eg \texttt{201507016}). It is significantly faster to decompose integers into date parts using arithmetic operations (\eg \texttt{mod} and \texttt{div}) than by using slower, locale-sensitive string parsing functions. Moreover, the number of possible formats is low enough that we may be able to enumerate them. Timestamp preparation from numeric representations is another form of preparation which we would like to automate.

In the course of our research, we have also identified a number of date part variants (\eg ordinal dates, four-letter month abbreviations, alternate meridian markers and postfix quarter syntax) that we would like to commend to the ICU project and maybe provide implementations of.

### 2015-07-07 intro.tex

In recent years, there has been a growth of interest in data visualization technologies for human-assisted data analysis using systems such as \cite{Stolte:2008,Ahlberg:1996}. While computers can provide high-speed and high-volume data processing, humans have domain knowledge and the ability to process data in parallel, notably by using our visual systems. Most importantly, humans provide the definition of what is valuable in an analysis. Accordingly, human/computer analytic systems are essential to extracting knowledge of value to humans and their societies from the large amounts of data being generated today.

@@

\subsection{The Tableau Ecosystem}
In this paper, we shall be looking at an example of this problem in the context of the Tableau system. Tableau is a commercial visual analysis environment derived from the Polaris~\cite{Stolte:2008} system developed at Stanford. In addition to an interactive desktop application for creating data visualizations, the Tableau ecosystem also includes servers for publishing and sharing interactive visualizations. These servers can be privately operated by individual customers or hosted in the cloud.

\subsubsection{Tableau Public}
Of particular relevance in the present work is a free version of the server environment called Tableau Public. Tableau Public (or ``Public'') allows authors to publish data visualizations that can be shared with the general public. The published data sets are limited to 100K rows and must be stored using the Tableau Data Engine (or ``TDE'')~\cite{Wesley:2011,Wesley:2014}, but are otherwise free to use the full analytic capabilities of the Tableau system. In particular, the TDE provides native support for all analytic calculations defined by the authoring and publishing components. Moreover, visualizations published to Public are uploaded as complete Tableau workbooks, including the data model developed as part of the analysis. This makes it a rich source of data on the analytic habits and frustrations of Tableau users.

\subsubsection{Functions}
While the existing standards for query languages are helpful for defining a core set of row-level functions, there are many useful functions beyond the standards that are only explicitly supported by a subset of RDBMSes, often with different names. And even when a database does not provide an implementation of a particular function, it is usually possible to generate it as an inline combination of existing functionality. Tableau's calculation language is designed to be a lingua franca that tames this Babel of dialects by providing a common syntax for as many of these functions as possible.

\subsection{DATEPARSE}
Among the recent additions to the Tableau function library is a function called \dateparse, the usability of which is the focus of the work in this paper.

@@

@@

The SQL-99 standard defines three temporal scalar types: \texttt{DATE}, \texttt{TIMESTAMP} and \texttt{TIME}. They are typically implemented as fixed-point types containing an offset from some epoch (\eg Julian Days.) This makes them compact to store and allows some temporal operations to be implemented very efficiently using arithmetic. Thus from the RDBMS perspective, representing dates in scalar form provides numerous benefits for users, both in terms of available operations and query performance.

Tableau models the first two of these types as ``Date'' and ``Date \& Time''; the third (pure time) is folded into Date \& Time by appending it to a fixed date of \texttt{1899-12-30}. From the analytic perspective, date types are dimensional (\ie independent variables) and can be used as either categorical (simply ordered) or quantitative (ordered with a distance metric) fields.

Categorical dates have a natural hierarchy associated with them generated by calendar binning. The visualization in Figure \ref{fig:I1} shows an example of a bar chart employing binned categorical dates in a year/quarter hierarchy.

@@

Quantitative dates are typically used for time series on an axis that maps the underlying distance measure to display pixel distance. The quantitative analog to categorical binning is truncation whereby all the dates in a bin are represented by the first date in the bin. Truncation accurately preserves the distance semantics while enabling roll-up to coarser levels of detail. The visualization in Figure \ref{fig:I2} shows the same sales data in a time series rolled up to the quarter level.

These types of visualizations are difficult to specify with simple categorical string representations of dates. Thus the conversion of unparsed strings to date scalars is an important component of data modeling for analytics.

\subsubsection{Parsing}
One of the oldest data preparation problems observed in Tableau is the parsing of date scalars so that users can perform these kinds of analyses. Training materials from the early days of the company included examples of how to convert columns of integers in the form \texttt{yyyyMMdd} to date scalars. The documented solution at the time was to convert the integer to a string and perform some locale-dependent string operations before casting the string back to a date. Unfortunately, this approach had a number of problems:
\begin{itemize}
\item String operations are notoriously slow compared to scalar operations (typically 10-100x slower)
\item Default parsing of date formats is locale-dependent, and may not work when the workbook is shared across an international organization (\eg between the US and European offices)
\item The expression code was hard to understand and maintain because it used a verbose, general-purpose string-handling syntax instead of a domain language.
\end{itemize}

This is but a single format. Our studies of the workbooks on Public suggest that there are hundreds of distinct temporal date formats in user data sets. Some are common, but others can be quite idiosyncratic. Table 1 shows a selection of unusual date formats found on Public. The first example shows a time zone in the middle of the date and a year after the time; the second shows a leading unmatched bracket and a colon between the date and time components; the third shows confusion between the seconds' decimal point and the time part delimiter; the fourth shows a two digit year apostrophe on a four digit year and the fifth shows a dash separating the date and time components.

@@

Our first solution to this problem was to add a new function to the Tableau calculation language called \dateparse. This function would take a string column and convert it to a datetime using a special purpose domain language for describing dates. Such functions exist in a number of databases (\eg Oracle, Postgres and MySQL) and adding it to the TDE was straightforward, so it was a natural addition to the function library.

We chose the date parsing syntax defined by the International Components for Unicode (or ICU) project~\cite{ICU} for the common domain language because it mirrored what was available in the Tableau code base, and translating this syntax into the formats used by various database vendors (\eg MySQL, Oracle, Postgres) was relatively painless. There are a few patches of non-overlapping functionality, but they tend to be obscure and we provide warnings when functionality is missing.

\subsubsection{	Usability}
After shipping this new function to our user base, we investigated how it was being applied in the field. We examined a few months of workbooks that had been uploaded to Public after the release to see if and how it was being used. We found that although there were a number of new calculations using \dateparse, the error rate for the domain language syntax was about 15\%. \dateparse was capable of solving the problem, and some users were able to discover it, but the syntax did not appear to be easy to use reliably.

\subsection{Automated Format Extraction}
One solution might have been to design a graphical environment that enabled users to construct valid patterns. This would have involved a substantial development investment with no guarantee that the result would be correct if the user misunderstood the environment. Instead, we have developed two algorithms for automating the derivation of the format string, each of which uses a different machine learning technique. Both algorithms result in over 90\% parsing accuracy -- and 100\% syntactic correctness because they are machine generated.

@@

\subsection{Related Work}
Data preparation has been a known analytic bottleneck since at least the description of the Potter's Wheel system~\cite{Raman:2001}. Since then, several other interactive data preparation systems have been proposed, including Data Wrangler~\cite{Kandel:2011} and Google refine~\cite{Refine}. While effective, these systems all make assumptions about possible date formats (\eg the domain system in~\cite{Raman:2001}), which we suggest are too restrictive for real world data.

The Minimum Descriptive Length technique was first proposed in Rissanen~\cite{Rissanen:1978}.  We have built on the version described in~\cite{Raman:2001}.

Related work on parsing languages, outlier detection.

>>>>

In recent years, there has been a growth of interest in data visualization technologies for human-assisted data analysis using systems such as Polaris~\cite{Stolte:2008} and Spotfire~\cite{Ahlberg:1996}. While computers can provide high-speed and high-volume data processing, humans have domain knowledge and the ability to process data in parallel, notably by using our visual systems. Most importantly, humans provide the definition of what is valuable in an analysis. Accordingly, human/computer analytic systems are essential to extracting knowledge of value to humans and their societies from the large amounts of data being generated today.

@@

\subsubsection{Functions}
While the existing standards for query languages are helpful for defining a core set of row-level functions, there are many useful functions beyond the standards that are only explicitly supported by a subset of RDBMSes, often with different names. And even when a database does not provide an implementation of a particular function, it is usually possible to generate it as an inline combination of existing functionality. The calculation languages of visualization systems can thus be designed to provide a \textit{lingua franca} that tames this Babel of dialects by providing a common syntax for as many of these functions as possible.

\subsection{\dateparse}
The focus of this work is the usability of one such function from our visualization system that we will call \dateparse.

@@

The SQL-99 standard defines three temporal scalar types: \texttt{DATE}, \texttt{TIMESTAMP} and \texttt{TIME}. They are typically implemented as fixed-point types containing an offset from some epoch (\eg Julian Days.) This makes them compact to store using column store coompression techniques such as those in C-Store~\cite{Stonebraker:2005} and MonetDB/X100~\cite{Zukowski:2006}, and further allows some temporal operations to be implemented very efficiently using simple arithmetic. Thus from the RDBMS perspective, representing dates in scalar form provides  benefits for users, both in terms of available operations and query performance.

@@

Our visualization system models the first two of these SQL-99 types natively; the third (pure time) is folded into \texttt{TIMESTAMP} by appending it to a fixed date of \texttt{1899-12-30} (a convention derived from Microsoft Excel). From the analytic perspective, date types are dimensional (\ie independent variables) and can be used as either \textit{categorical} (simply ordered) or \textit{quantitative} (ordered with a distance metric) fields.

Categorical dates have a natural hierarchy associated with them generated by calendar \textit{binning}. The visualization in Figure \ref{fig:I1} shows an example of a bar chart employing binned categorical dates in a year/quarter hierarchy.

@@

Quantitative dates are typically used for time series on an axis that maps the underlying distance measure to display pixel distance. The quantitative analog to categorical binning is \textit{truncation} whereby all the dates in a bin are represented by the first date in the bin. Truncation accurately preserves the distance semantics while enabling roll-up to coarser levels of detail. The visualization in Figure \ref{fig:I2} shows the same sales data in a time series rolled up to the quarter level.

These types of visualizations are difficult to specify with simple categorical string representations of dates. Thus the conversion of unparsed strings to date scalars is also an important component of data modeling for analytics.

\subsubsection{Parsing}
This parsing of date representations is one of the oldest data preparation problems we have observed. A common version of this problem is how to convert columns of integers of the form \texttt{yyyyMMdd} to date scalars. Naive users often solve this problem by converting the integer to a string and performing some locale-dependent string operations before casting the string back to a date. Unfortunately, this approach has a number of problems:
\begin{itemize}
\item String operations are notoriously slow compared to scalar operations (typically 10-100x slower in modern RDBMSes)
\item Default parsing of date formats is locale-dependent, and may not work when the analysis is shared across an international organization (\eg between the US and European offices)
\item The parsing code was hard to understand and maintain because it used a verbose, general-purpose string-handling syntax instead of a specialized domain language.
\end{itemize}

And this is but a single date format. Our studies of online data collections suggest that there are hundreds of distinct temporal date formats in user data sets. Some are common, but others can be quite idiosyncratic. Table 1 shows a selection of unusual date formats found in our test data. The first example shows a time zone in the middle of the date and a year after the time; the second shows a leading unmatched bracket and a colon between the date and time components; the third shows confusion between the seconds' decimal point and the time part delimiter; the fourth shows a two digit year apostrophe on a four digit year and the fifth shows a dash separating the date and time components.

@@

Our first attempt to help users with this problem was to add a new function to the the calculation language called \dateparse. This function would take a string column and convert it to a datetime using a special purpose domain language for describing dates. Such functions exist in a number of databases (\eg MySQL, Oracle and Postgres), so it was a natural addition to the function library.

We chose the date parsing syntax defined by the International Components for Unicode (or ICU) project~\cite{ICU} for the common domain language because it mirrored what was already available in our code base, and translating this syntax into the formats used by various database vendors was relatively painless. There are a few patches of non-overlapping functionality, but they tend to be obscure and we provide warnings when functionality is missing.

\subsubsection{	Usability}
After providing this new function to our user base, we waited a few months and then investigated how it was being applied in a free, cloud-based version of the system. We found that although there were a number of new calculations using \dateparse, the error rate for the domain language syntax was about 15\%. \dateparse was capable of solving the problem, and some users were able to discover it, but the syntax did not appear to be easy to use reliably.

\subsection{Automated Format Extraction}
One solution to this problem might have been to design a graphical environment that enabled users to construct valid patterns. This would have involved a substantial development effort with no guarantee that the result would be correct if the user misunderstood the environment. Instead, we have developed two algorithms for automating the derivation of the format string, each of which uses a different machine learning technique. Both algorithms result in over 90\% parsing accuracy -- and 100\% syntactic correctness because they are machine generated.

@@

\subsection{Related Work}
Data preparation has been considered an analytic bottleneck since at least the description of the Potter's Wheel system~\cite{Raman:2001}. Since then, several other interactive data preparation systems have been proposed, including Data Wrangler~\cite{Kandel:2011} and Google Refine~\cite{Refine}. While effective, these systems all make assumptions about possible date formats, which we suggest are too restrictive for real world data.

Various approaches have been described for deriving regular expressions, and a good overview is provided by Li et al. in their paper on the ReLIE system for deriving regular expressions given a starting expression provided by a domain expert~\cite{Li:2008}. The Minimum Descriptive Length technique first described in Rissanen~\cite{Rissanen:1978} was used in~\cite{Raman:2001} to generate regular expressions.

Related work on parsing languages, outlier detection??

### 2015-07-07 mdl.tex

Knowing the locale of a string is not always helpful. In our data set, we found numerous cases where the locale was specified (\eg Sweden) but the data was actually in English. Accordingly, we test both locales and rank the combined results.
Tableau currently does not support non-Gregorian calendars. Therefore, we have built this system for the Gregorian calendar only. ICU does support non-Gregorian calendars and we expect this system could be applied to them as well.

>>>>

Knowing the locale of a string is not always helpful. In our data set, we found numerous cases where the locale was specified (\eg Sweden) but the data was actually in English. Accordingly, we test both locales and rank the combined results.

We have built this system for the Gregorian calendar only as our visualization system does not support non-Gregorian calendars and we have little evidence of other calendars (\eg Hebrew, Islamic Civil) being used for analytics. ICU supports non-Gregorian calendars and we expect the algorithm could be extended to them as well if needed.

### 2015-07-07 parameters.tex

\subsection{Input Data}
The training data and test data are taken from Public data sets. We selected a large number of string columns whose names suggested that they might contain temporal data. We included words from multiple languages besides English and extracted the column's locale (actually the collation) along with the data.

@@

\subsection{The ICU Date Format Language}
The date format syntax we selected is the one provided by the open-source project. We chose it because we were already using ICU in the code base, we had access to the source code and it provides localized date part data for a large number of languages.

@@

ICU's date parsing APIs allow the specification of default values for parts when a format does not contain them. In our implementation, all time fields are set to \texttt{0} (midnight) and the date fields are set based on whether the format contains any date part specifications. When date parts are present, we use \texttt{2000-01-01} as the set of default date parts as it is the start of a leap year. When dealing with pure time formats, we use \texttt{1899-12-30}, (which is the date used internally by Tableau to signal pure time values.)

ICU will also parse Time Zones (which we recognize but do not use in Tableau) and Quarters (which we interpret as the first month of the period.) RDBMSes such as Oracle and Postgres that support time zones will be able to take advantage of the generated time zone field.

>>>>

\subsection{Input Data}
The training data and test data are taken from data sets published to our free, online visualization service. The published data includes both the raw data and how it was used for analysis. The service limits data sets to a maximum of 100K rows stored in a database which supports an implementation of \dateparse.  We selected a large number of string columns whose names suggested that they might contain temporal data. We included words from multiple languages besides English and extracted the column's locale (actually the collation) along with the data.

@@

\subsection{The ICU Date Format Language}
The date format syntax we selected is the one provided by the ICU open-source project. We chose it because we were already using ICU in the code base, we had access to the source code, and it provides localized date part data for a large number of languages.

@@

ICU's date parsing APIs allow the specification of default values for parts when a format does not contain them. In our implementation, all time fields are set to \texttt{0} (midnight) and the date fields are set based on whether the format contains any date part specifications. When date parts are present, we use \texttt{2000-01-01} as the set of default date parts as it is the start of a leap year. When dealing with pure time formats, we model the output as a date/time and use \texttt{1899-12-30} for the date parts.

ICU will also parse Time Zones (which the algorithms recognize but the visualization system does not model) and Quarters (which we interpret as the first month of the period.) RDBMSes such as Oracle and Postgres that support time zones will be able to take advantage of  identified time zone fields.
